Regime-Specific Performance of 1D CNN and FCNN Architectures for Non-linear Matter Power Spectrum Emulation in LCDM Cosmology

### 2025-05-02 Project4/paper_v2.tex
fix: Correct LaTeX errors for compilation and improve formatting

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}
\usepackage{amsmath}
\usepackage{multirow}

@@

We generated a training and testing dataset of non-linear matter power spectra using the \texttt{classy\_sz} Boltzmann solver. To efficiently explore the cosmological parameter space, we employed Latin Hypercube Sampling (LHS) to create 500,000 unique samples. Each sample represents a point in the 6-dimensional parameter space defined by the following cosmological parameters and redshift: baryon density ($\omega_b$), cold dark matter density ($\omega_{cdm}$), Hubble constant ($H_0$), the amplitude of the primordial power spectrum ($\log A$), spectral index ($n_s$), and redshift ($z$). The sampling ranges for each parameter were: $\omega_b \in [0.01933, 0.02533]$, $\omega_{cdm} \in [0.08, 0.20]$, $H_0 \in [40, 100]$ km/s/Mpc, $\log A \in [2.5, 3.5]$, $n_s \in [0.8, 1.2]$, and $z \in [0, 1]$. The LHS technique ensures a quasi-random distribution of samples that efficiently covers the entire parameter space, minimizing clustering and maximizing coverage. Each sample is represented as a vector $(\omega_b, \omega_{cdm}, H_0, \log A, n_s, z)$.

@@

For each of the 500,000 parameter sets generated via LHS, we used the \texttt{classy\_sz} code to calculate the non-linear matter power spectrum, $P(k, z)$. \texttt{classy\_sz} was configured to output the power spectrum over a pre-defined $k$-grid, spanning the range $k \in [10^{-4}, 10]$ $h$/Mpc. This range encompasses the linear, quasi-linear, and non-linear regimes of structure formation. The $k$-grid was held constant across all simulations to ensure consistency and facilitate direct comparison of the resulting power spectra. The specific $k$ values were determined by the default configuration of \texttt{classy\_sz}. The output of \texttt{classy\_sz} for each parameter set is a one-dimensional array of $P(k, z)$ values corresponding to the fixed $k$-grid.

@@

The generated data was structured into a dataset suitable for training and testing neural network emulators. Each data point consisted of a 6-dimensional input vector (the cosmological parameters and redshift: $\omega_b, \omega_{cdm}, H_0, \log A, n_s, z$) and a corresponding one-dimensional output array representing the non-linear matter power spectrum $P(k, z)$ at the pre-defined $k$ values. The complete dataset was then randomly split into training and testing sets, with 80\% of the data allocated to the training set (400,000 samples) and 20\% allocated to the testing set (100,000 samples). This split was performed using a stratified random sampling approach to ensure that the distribution of cosmological parameters and redshifts in the training and testing sets was representative of the overall distribution. This helps to avoid biases in the emulator's performance evaluation.

@@

This section presents a detailed analysis of the performance of 1D CNN and FCNN architectures for emulating the non-linear matter power spectrum, $P(k, z)$, within the $\Lambda$CDM cosmological framework. The emulators were trained and tested on a dataset of 500,000 spectra generated using a Latin Hypercube sampling of cosmological parameters and redshifts ($z \in [0, 1]$), calculated using the \texttt{classy\_sz} Boltzmann solver. The evaluation focuses on accuracy, measured by Mean Squared Error (MSE) and relative error, as well as computational efficiency. We analyze the results across distinct physically motivated regimes of wavenumber ($k$), redshift ($z$), and cosmological parameter space. Figures \ref{fig:true_vs_emulated_cnn1d} and \ref{fig:true_vs_emulated_fcnn} show a comparison of the true and emulated power spectra for the 1D CNN and FCNN models, respectively, demonstrating the overall agreement between the models and the true spectra.

@@

We assessed the performance of both architectures using the test set, comprising 50,000 spectra. The primary metrics were MSE, a measure of the average squared difference between the predicted and true power spectra, and the median absolute relative error, $|\Delta P/P|$, providing a robust measure of fractional deviation. The $k$-space was divided into three regimes: linear ($k < 0.1\,h/\mathrm{Mpc}$), quasi-linear ($0.1 \leq k < 0.5\,h/\mathrm{Mpc}$), and non-linear ($k \geq 0.5\,h/\mathrm{Mpc}$).

@@

Figures \ref{fig:error_vs_k_cnn1d} and \ref{fig:error_vs_k_fcnn} show the relative error as a function of wavenumber $k$ for the 1D CNN and FCNN, respectively. As shown in Figure \ref{fig:error_vs_k_cnn1d}, the 1D CNN exhibits a median relative error that oscillates significantly around $\log_{10} k \sim -1$. Figure \ref{fig:error_vs_k_fcnn} shows that the FCNN's relative error increases significantly at high $k$, with oscillations around $k \sim 10^{-1} \, h/\mathrm{Mpc}$.

@@

The median and 68\% interval of the relative error ($\Delta P/P$) as a function of $k$ reveal distinct performance characteristics across the different regimes. In the linear regime ($k < 0.1\,h/\mathrm{Mpc}$), both models achieve sub-percent median errors, with the 1D CNN exhibiting slightly better performance than the FCNN. As $k$ increases into the quasi-linear regime ($0.1 \leq k < 0.5\,h/\mathrm{Mpc}$), the errors increase, but the 1D CNN maintains a lower median and narrower spread compared to the FCNN. In the non-linear regime ($k \geq 0.5\,h/\mathrm{Mpc}$), both models show a further increase in error, but the 1D CNN remains more robust, with the FCNN exhibiting a broader error distribution.

@@

Figures \ref{fig:error_vs_z_cnn1d} and \ref{fig:error_vs_z_fcnn} illustrate the relative error as a function of redshift for the 1D CNN and FCNN, respectively, across different $k$-regimes. The 1D CNN (Figure \ref{fig:error_vs_z_cnn1d}) shows a decreasing trend with increasing redshift in the linear regime, a slightly increasing trend in the quasi-linear regime, and a decreasing trend in the non-linear regime. The FCNN (Figure \ref{fig:error_vs_z_fcnn}) exhibits the smallest relative error in the linear regime, consistently below 0.003 across all redshifts, while the quasi-linear regime shows the largest errors, peaking around $z=0.6$.

@@

The results demonstrate that both the 1D CNN and FCNN architectures can effectively emulate the matter power spectrum. However, the 1D CNN consistently outperforms the FCNN, especially in the quasi-linear and non-linear regimes. This is evident from the error analysis presented in Figures \ref{fig:error_vs_k_cnn1d}, \ref{fig:error_vs_k_fcnn}, \ref{fig:error_vs_z_cnn1d}, and \ref{fig:error_vs_z_fcnn}. This suggests that the convolutional layers in the 1D CNN architecture are better suited for capturing the complex, scale-dependent features of the non-linear matter power spectrum. The ability to capture local correlations in $P(k)$ is particularly important at high $k$. The inference speed of both models is comparable, making the 1D CNN the preferred choice for applications requiring high-fidelity emulation of $P(k)$ across all regimes. The slightly higher complexity of the 1D CNN does not introduce a significant computational overhead during inference.

>>>>

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}
\usepackage{amsmath}
\usepackage{graphicx} % Added for \includegraphics
\usepackage{multirow}

@@

We generated a training and testing dataset of non-linear matter power spectra using the \texttt{classy\_sz} Boltzmann solver. To efficiently explore the cosmological parameter space, we employed Latin Hypercube Sampling (LHS) to create 500,000 unique samples. Each sample represents a point in the 6-dimensional parameter space defined by the following cosmological parameters and redshift: baryon density ($\omega_b$), cold dark matter density ($\omega_{cdm}$), Hubble constant ($H_0$), the amplitude of the primordial power spectrum ($\log_{10} A$), spectral index ($n_s$), and redshift ($z$). The sampling ranges for each parameter were: $\omega_b \in [0.01933, 0.02533]$, $\omega_{cdm} \in [0.08, 0.20]$, $H_0 \in [40, 100]\,\mathrm{km}\,\mathrm{s}^{-1}\,\mathrm{Mpc}^{-1}$, $\log_{10} A \in [2.5, 3.5]$, $n_s \in [0.8, 1.2]$, and $z \in [0, 1]$. The LHS technique ensures a quasi-random distribution of samples that efficiently covers the entire parameter space, minimizing clustering and maximizing coverage. Each sample is represented as a vector $(\omega_b, \omega_{cdm}, H_0, \log_{10} A, n_s, z)$.

@@

For each of the 500,000 parameter sets generated via LHS, we used the \texttt{classy\_sz} code to calculate the non-linear matter power spectrum, $P(k, z)$. \texttt{classy\_sz} was configured to output the power spectrum over a pre-defined $k$-grid, spanning the range $k \in [10^{-4}, 10]\,h\,\mathrm{Mpc}^{-1}$. This range encompasses the linear, quasi-linear, and non-linear regimes of structure formation. The $k$-grid was held constant across all simulations to ensure consistency and facilitate direct comparison of the resulting power spectra. The specific $k$ values were determined by the default configuration of \texttt{classy\_sz}. The output of \texttt{classy\_sz} for each parameter set is a one-dimensional array of $P(k, z)$ values corresponding to the fixed $k$-grid.

@@

The generated data was structured into a dataset suitable for training and testing neural network emulators. Each data point consisted of a 6-dimensional input vector (the cosmological parameters and redshift: $\omega_b, \omega_{cdm}, H_0, \log_{10} A, n_s, z$) and a corresponding one-dimensional output array representing the non-linear matter power spectrum $P(k, z)$ at the pre-defined $k$ values. The complete dataset was then randomly split into training and testing sets, with 80\% of the data allocated to the training set (400,000 samples) and 20\% allocated to the testing set (100,000 samples). This split was performed using a stratified random sampling approach to ensure that the distribution of cosmological parameters and redshifts in the training and testing sets was representative of the overall distribution. This helps to avoid biases in the emulator's performance evaluation.

@@

This section presents a detailed analysis of the performance of 1D CNN and FCNN architectures for emulating the non-linear matter power spectrum, $P(k, z)$, within the $\Lambda$CDM cosmological framework. The emulators were trained and tested on a dataset of 500,000 spectra generated using a Latin Hypercube sampling of cosmological parameters and redshifts ($z \in [0, 1]$), calculated using the \texttt{classy\_sz} Boltzmann solver. The evaluation focuses on accuracy, measured by Mean Squared Error (MSE) and relative error, as well as computational efficiency. We analyze the results across distinct physically motivated regimes of wavenumber ($k$), redshift ($z$), and cosmological parameter space. Figures~\ref{fig:true_vs_emulated_cnn1d} and \ref{fig:true_vs_emulated_fcnn} show a comparison of the true and emulated power spectra for the 1D CNN and FCNN models, respectively, demonstrating the overall agreement between the models and the true spectra.

@@

We assessed the performance of both architectures using the test set, comprising 50,000 spectra. The primary metrics were MSE, a measure of the average squared difference between the predicted and true power spectra, and the median absolute relative error, $|\Delta P/P|$, providing a robust measure of fractional deviation. The $k$-space was divided into three regimes: linear ($k < 0.1\,h\,\mathrm{Mpc}^{-1}$), quasi-linear ($0.1 \leq k < 0.5\,h\,\mathrm{Mpc}^{-1}$), and non-linear ($k \geq 0.5\,h\,\mathrm{Mpc}^{-1}$).

@@

Figures~\ref{fig:error_vs_k_cnn1d} and \ref{fig:error_vs_k_fcnn} show the relative error as a function of wavenumber $k$ for the 1D CNN and FCNN, respectively. As shown in Figure~\ref{fig:error_vs_k_cnn1d}, the 1D CNN exhibits a median relative error that oscillates significantly around $\log_{10} k \sim -1$. Figure~\ref{fig:error_vs_k_fcnn} shows that the FCNN's relative error increases significantly at high $k$, with oscillations around $k \sim 10^{-1}\,h\,\mathrm{Mpc}^{-1}$.

@@

The median and 68\% interval of the relative error ($\Delta P/P$) as a function of $k$ reveal distinct performance characteristics across the different regimes. In the linear regime ($k < 0.1\,h\,\mathrm{Mpc}^{-1}$), both models achieve sub-percent median errors, with the 1D CNN exhibiting slightly better performance than the FCNN. As $k$ increases into the quasi-linear regime ($0.1 \leq k < 0.5\,h\,\mathrm{Mpc}^{-1}$), the errors increase, but the 1D CNN maintains a lower median and narrower spread compared to the FCNN. In the non-linear regime ($k \geq 0.5\,h\,\mathrm{Mpc}^{-1}$), both models show a further increase in error, but the 1D CNN remains more robust, with the FCNN exhibiting a broader error distribution.

@@

Figures~\ref{fig:error_vs_z_cnn1d} and \ref{fig:error_vs_z_fcnn} illustrate the relative error as a function of redshift for the 1D CNN and FCNN, respectively, across different $k$-regimes. The 1D CNN (Figure~\ref{fig:error_vs_z_cnn1d}) shows a decreasing trend with increasing redshift in the linear regime, a slightly increasing trend in the quasi-linear regime, and a decreasing trend in the non-linear regime. The FCNN (Figure~\ref{fig:error_vs_z_fcnn}) exhibits the smallest relative error in the linear regime, consistently below 0.003 across all redshifts, while the quasi-linear regime shows the largest errors, peaking around $z=0.6$.

@@

The results demonstrate that both the 1D CNN and FCNN architectures can effectively emulate the matter power spectrum. However, the 1D CNN consistently outperforms the FCNN, especially in the quasi-linear and non-linear regimes. This is evident from the error analysis presented in Figures~\ref{fig:error_vs_k_cnn1d}, \ref{fig:error_vs_k_fcnn}, \ref{fig:error_vs_z_cnn1d}, and \ref{fig:error_vs_z_fcnn}. This suggests that the convolutional layers in the 1D CNN architecture are better suited for capturing the complex, scale-dependent features of the non-linear matter power spectrum. The ability to capture local correlations in $P(k)$ is particularly important at high $k$. The inference speed of both models is comparable, making the 1D CNN the preferred choice for applications requiring high-fidelity emulation of $P(k)$ across all regimes. The slightly higher complexity of the 1D CNN does not introduce a significant computational overhead during inference.
